\documentclass[a4paper,twocolumn,10pt]{article}
\input{preamble}

\title{Notes for M.A.~Armstrong's \textit{Groups and Symmetry}}
\author{Christian Stigen Larsen}
\date{March 2016}

\begin{document}
  \maketitle
  \section{Symmetries of the Tetrahedron}
  \paragraph{Symmetry group} Captures the rules of how symmetries combine for a
  given object.

  \paragraph{Order of operations} In the \textit{product}\footnote{Rotations,
  flips, multiplications, additions, etc. Same order as functional
  composition.} $xyz$, do $z$ first, then $y$ and finally $x$. If order doesn't
  matter in $G$, it's commutative (or \textbf{\textit{abelian}}). Remember to
  label geometric vertices.

  \section{Axioms}
  \paragraph{Group}  Set $G$ with \textit{multiplication} (addition,
  rotation, etc.) satisfying
  \begin{itemize}
    \item \textbf{\textit{associativity}}, i.e.~$(xy)z = x(yz)$
    \item \textbf{\textit{identity element}} $e \in G$ such that $xe=x=ex$
    \item \textbf{\textit{inverse}} $e \in G$ such that $x^{-1}x=e=xx^{-1}$
  \end{itemize}

  \paragraph{Properties common to all groups}
  \begin{itemize}
    \item The identify element of a group is unique.
    \item The inverse of each element of a group is unique.
  \end{itemize}

  \section{Numbers}
  \paragraph{Addition of $\mathbb{Z}, \mathbb{Q}, \mathbb{R}, \mathbb{C}$}
  \begin{itemize}
    \item Identity is zero
    \item $-x$ is the inverse
  \end{itemize}
  \paragraph{Multiplication}
  \begin{itemize}
  \item For $\mathbb{Q}-\{0\}$, $\mathbb{Q}^{\textbf{pos}}$,
    $\mathbb{R}-\{0\}$, $\mathbb{R}^{\textbf{pos}}$, $\{+1,-1\}$,
    $\mathbb{C}-\{0\}$, $\mathcal{C}$\footnote{Complex numbers of modulus 1.},
    $\{\pm 1, \pm i\}$: $e=1$ and $x^{-1}=\nicefrac{1}{x}$.
  \end{itemize}

  \paragraph{$\mathbb{Z}$ under addition modulus $n$} $e=0$,
  $x^{-1}=n-x$ for $x\ne0$, finite \textit{abelian} group and denoted
  $\mathbb{Z}_n$.

  \paragraph{$\mathbb{Z}$ under multiplication modulus $n$} Requires $n$ to be
  prime.

  \section{Dihedral Groups}
  When $n\geqslant3$ we can manufacture a plate whish has $n$ equal sides. These are
  the non-commutative \textit{dihedral rotational symmetry groups} $D_n$. E.g.
  $D_3 = \{e,r,r^2,s,rs,r^2s\}$. $x^mx^n=x^{m+n}$ and $(x^m)^n=x^{mn}$ provided
  we interpret $x^0=e$. For any multiplication table, each element in $G$
  appears only once in every given column or row.

  $r^n=e$, $s^2=e$, $sr=r^{n-1}s$, $r^{n-1}=r^{-1}$, etc.

  Each element is of form $r^a$, $r^as$ where $0\leqslant a\leqslant n-1$.

  For $k=a+_nb$, $r^ar^b=r^k$ and $r^a(r^bs)=r^ks$.  For $l=a+_n(n-b)$,
  $(r^as)r^b=r^ls$ and $(r^as)(r^bs)=r^l$ --- thus $r$ and $s$
  \textbf{\textit{generate}} $D_n$.

  The \textbf{\textit{order}} $|G|$ is the number of elements in the group. If
  $x^n=e$, then the \textit{element} $x$ has \textit{finite} order $n$ when $n$
  is the smallest such $n$.

  \section{Subgroups and Generators}
  A \textbf{\textit{subgroup}} of $G$ is a subset of $G$ which itself forms a
  group under the multiplication of $G$. For $H$ to be a subgroup of $G$,
  $H<G$:
  \begin{itemize}
    \item $xy \in G$ for any $x,y \in H$
    \item $e_H \in G$
    \item For any $x \in H$, $x^{-1} \in G$
    \item Associativity in $G$ implies the same for $H$.
  \end{itemize}

  \paragraph{Subgroup generated by $x$, or $\langle x \rangle$}
  For an element $x$ in $G$, the set of all $x^n$ is a subgroup of $G$
  (remember $x^0=e$). Finite order $m$ means $x^0=e, x^1, \ldots, x^{m-1}$.
  So order of $x\in G$ is precisely the order of $\langle x\rangle$. If
  $\langle x\rangle=G$, i.e., generates all of $G$, then $G$ is a
  \textbf{\textit{cyclic group}}.

  \paragraph{Subgroup generated by $X$} If $X<G$\footnote{$X$ is a subgroup of
  $G$.} and, for example, $r$,$s$,$r^2$,$sr$ (called \textit{words} of $X$).

  \paragraph{Theorems}
  \begin{itemize}[leftmargin=0.45in]
    \item[\theorem{5.1}] A non-empty subset $H$ of a group $G$ is a subgroup of
      $G$ if and only if $xy^{-1}$ belongs to $H$ whenever $x$ and $y$ belong
      to $H$.

    \item[\theorem{5.2}] The intersection of two subgroups of a group is itself
      a subgroup.

    \item[\theorem{5.3}] Every subgroup of $\mathbb{Z}$ is cyclic. Every
      subgroup of a cyclic group is cyclic.
  \end{itemize}

  \section{Permutations}
  A \textit{permutation} is a bijection\footnote{A one-to-one mapping between
  the elements of two sets, meaning you can always go backwards as well.} from a
  set $X$ to itself (e.g.,~replace all $3$s with $1$s).  The collecticon of
  \textit{all} permutations of $X$ forms a group $S_x$ under composition of
  functions (who each perform one specific permutation). When $X$ consists of
  the first $n$ positive integers, we get the \textbf{\textit{symmetric group}}
  $S_n$ of degree $n$ and order $n!$. $S_3$ is not abelian

  $(a_1a_2\ldots a_k)$ is called a \textbf{\textit{cyclic permutation}},
  sending $a_1$ to $a_2$, $\ldots$ , $a_k$ to $a_1$. Its length is $k$ and a
  cyclic permutation of length $k$ is called a \textbf{\textit{k-cycle}}. A
  2-cycle is called a \textbf{\textit{transposition}}. Every element of $S_n$
  can be written as many such \textbf{\textit{disjoint}}, meaning no integer is
  moved by more than one of them. Therefore they are \textit{commutative}.

  \paragraph{A few tricks}
  \begin{itemize}
    \item Each \textit{element} of $S_n$ can be written as a product of cyclic
      permutations, and any cyclic permutation can be written as a product of
      transpositions: $(a_1a_2 \ldots a_k) = (a_1a_k) \ldots (a_1a_3)(a_1)(a_2)$.
      Therefore, each \textit{element} of $S_n$ can be written as a product of
      transpositions.
    \item $(ab) = (1a)(1b)(1a)$
    \item $(1k) = (k-1,k)$$\ldots$$(34) (23) (12) (23) (34)$$\ldots$$(k-1,k)$
  \end{itemize}

  \paragraph{Theorems}
  \begin{itemize}[leftmargin=0.45in]
    \item[\theorem{6.1}] The transpositions in $S_n$ together generate
      $S_n$.

    \item[\theorem{6.2a}] The transpositions $(12), (13), \ldots, ({1n)}$
      together generate $S_n$.

    \item[\theorem{6.2b}] The transpositions $(12), (23), \ldots, (n-1,n)$
      together generate $S_n$.

    \item[\theorem{6.3}] The transposition $(12)$ and the $n$-cycle $(12
      \ldots n)$ together generate $S_n$.
  \end{itemize}

  Any \textit{element} $\alpha$ of $S_n$ can be written as a product of
  \textit{transpositions} in many different ways. But the number of
  transpositions is always even or always odd. If $\alpha$ \textit{can} be
  written as the product of an even number of transpositions, then its sign
  must be $+1$; for odd, it is $-1$. Therefore, by the first trick above, a
  \textit{cyclic permutation} is even precisely when its length is odd.

  \paragraph{Theorems}
  \begin{itemize}[leftmargin=0.45in]
    \item[\theorem{6.4}] The even permutations in $S_n$ form a subgroup of
      order $n!/2$ called the \textbf{\textit{alternating group $A_n$}} of
      degree $n$.

    \item[\theorem{6.5}] For $n\geqslant 3$ the 3-cycles generate $A_n$.
  \end{itemize}

  \section{Isomorphisms}
  If two multiplication tables have corresponding elements and products, they
  are \textbf{\textit{isomorphic}}.

  Two groups $G$ and $G'$ are \textbf{\textit{isomorphic}} if there is a
  bijection $\phi$ from $G$ to $G'$ which satisfies $\phi(xy) =
  \phi{(x)}\phi{(y)}$ for all $x,y \in G$. The function $\phi$ is called an
  \textbf{\textit{isomorphism}} between $G$ and $G'$.
  This is written $G \cong G'$.

  \paragraph{Notes}
  \begin{itemize}
    \item $G$ and $G'$ have the same order.
    \item $\phi{(x)}^{-1} = \phi{(x^{-1})}$ for all $x \in G$.
    \item If $G$ is abelian, then so is $G'$.
    \item If $H$ is a subgroup of $G$ then $\phi{(H)}$ a subgroup of $G'$.
    \item An isomorphism preserves the order of each element.
    \item If $\phi\colon G \rightarrow G'$ and $\psi\colon G' \rightarrow G''$ are both
      isomorphisms, then the composition $\psi\phi\colon G \rightarrow G''$ is also
      an isomorphism.
  \end{itemize}
  \paragraph{Examples}
  \begin{itemize}
    \item $\phi\colon \mathbb{R} \rightarrow \mathbb{R}^{\textbf{pos}}$ by
      $\phi{(x)} = e^x$ and $\phi{(x+y)}=e^{x+y}=e^xe^y=\phi{(x)}\phi{(y)}$.
    \item The non-abelian, rotational group $G$ for the tetrahedron is isomorphic to $A_4$.
    \item Any infinite cyclic group $G$ is isomorphic to $\mathbb{Z}$ by
      $\phi{(x^m)} = m$ and $\phi{(x^mx^n)} =
      \phi{(x^{m+n})}=m+n=\phi{(x^m)}+\phi{(x^n)}$.
    \item Any finite cyclic group of order $n$ is isomorphic to $\mathbb{Z}_n$
      by $\phi{(x^m)} = m\Mod{n}$.
    \item The numbers $1$, $-1$, $i$, $-i$ form a group under complex
      multiplication. It is cyclic, and $i$, $-i$ are both generators. It gives
      two isomorphisms between this group and $\mathbb{Z}_4$.
    \item $D_3$ and $S_3$ are isomorphic.
    \item There is no isomorphism between $\mathbb{Q}$ and
      $\mathbb{Q}^{\textbf{pos}}$.
  \end{itemize}

  \section{Plato's Solids and Cayley's Theorem}
  \textit{Remember:} A surjection between two finite sets which have the same
  number of elements must be a bijection.

  \begin{itemize}
    \item The rotational symmetry group of the tetrahedron is isomorphic to
      $A_4$.

    \item The cube and octehedron both have rotational symmetry groups which
      are isomorphic to $S_4$.

    \item The dodecahedron and icosahedron both have rotational symmetry groups
      which are isomorphic to $A_5$.

    \item If two solids are \textbf{\textit{dual}} to one another, their
      rotational symmetry groups are isomorphic.
  \end{itemize}

  \paragraph{Theorems}
  \textit{Every} group is isomorphic to a subgroup of permutations:
  \begin{itemize}[leftmargin=0.45in]
    \item[\theorem{8.1}] \textbf{Cayley's Theorem.} Let $G$ be a group, then
      $G$ is isomorphic to a subgroup of $S_G$.

    \item[\theorem{8.2}] If $G$ is a finite group of order $n$, then $G$ is
      isomorphic to a subgroup of $S_n$.
  \end{itemize}

  \section{Matrix Groups}
  The set of all invertible $n \times n$ matrices with real numbers as entries
  forms a group under matrix multiplication: Matrix multiplication is
  associative, the $n \times n$ identity matrix $I_n=\epsilon$ and the inverse
  of $AB$ is $B^{-1}A^{-1}$. This group is called the \textbf{\textit{General
  Linear Group}}, $GL_n$.

  Matrix multiplication is not commutative for $n \geqslant 2$, so we have a
  family of \textit{infinite non-abelian} groups $GL_2$, $GL_3$, etc. For $n=1$
  the single entry must be a non-zero number (the matrix is invertible), and
  reduces to ordinary multiplication of numbers. Hence, $GL_1 \cong
  \mathbb{R}-\{0\}$.

  $AB^{-1}$ is orthogonal and by theorem (\ref{theorem-5.1}.1) the collection
  of all $n \times n$ orthogonal matrices is a subgroup of $GL_n$. This
  subgroup is called the \textbf{\textit{Orthogonal Group}}, $O_n$. Those
  elements of $O_n$ which have determinant equal to $+1$ form a subgroup of
  $O_n$ called the \textbf{\textit{Special Orthogonal Group}}, $SO_n$.

  \textit{No further notes here, at the moment.}

  \section{Products}
  The \textbf{\textit{direct product}} $G \times H$ of two groups $G$ and $H$
  is constructed by $(g,h)(g',h') = (gg',hh')$, where $g, g' \in G$ and $h, h'
  \in H$. Thus, $(gg',hh') \in G \times H$ and $G \times H$ is a group. The
  correspondence $(g,h) \rightarrow (h,g)$ means that $G \times H$ is
  isomorphic to $H \times G$. Unless either of $G$ or $H$ are of infinite
  order, $|G \times H| = |G|\cdot|H|$. If both $G$ and $H$ are abelian, so is
  $G \times H$. In reverse, if $G \times H$ is abelian, so are both $G$ and
  $H$.

  E.g., the elements of $\mathbb{Z}_2 \times \mathbb{Z}_3$ are
  $\{0,1\}\times\{0,1,2\} = \{(0,0), (0,1), (0,2), (1,0), (1,1), (1,2)\}$ and
  their elements are combined by $(x,y) + (x',y') = (x \,+_2\, x', y \,+_3\, y')$.
  We follow the convention of using $+$ for the group structure whenever we
  have products of cyclic groups. As continually adding $(1,1)$ to itself, we
  can fill out the whole group, and therefore $\mathbb{Z}_2 \times
  \mathbb{Z}_3$ is cyclic and isomorphic to $\mathbb{Z}_6$.

  \paragraph{Klein's group} $\mathbb{Z}_2 \times \mathbb{Z}_2$ is non-cyclic
  and isomorphic to the group of plane symmetries of a chessboard.

  We write $\mathbb{R}^n$ for the direct product of $n$ copies of $\mathbb{R}$.

  \paragraph{Theorem \theorem{10.1}} $\mathbb{Z}_m \times \mathbb{Z}_n$ is
  cyclic if and only if the highest common factor of $m$ and $n$ is 1.

  \paragraph{Theorem \theorem{10.2}} If $H$ and $K$ are subgroups of $G$ for
  which $HK=G$, if they have only the identity element in common, and if every
  element of $H$ commutes with every element of $K$, then $G$ is isomorphic to
  $H \times K$.

  The linear transformation $f_J\colon \mathbb{R}^3 \rightarrow \mathbb{R}^3$
  sends each vector $\bm{x}$ to $\bm{-x}$ and is called
  \textbf{\textit{central inversion}}.

  \textit{Some important notions at the end of the chapter have been left out,
  currently.}

  \section{Lagrange's Theorem}
  Let $H<G$ and break it up as the union of the $k+1$ pieces $H, g_1H, \ldots,
  g_kH$, then $|G|=(k+1)|H|$.

  \theorem{11.1} The order of a subgroup of a finite group is always a divisor
  of the order of the group.

  \textbf{Note:} The opposite is not true; the existence of a divisor $m$
  of $|G|$ does \textit{not} imply the existence of a subgroup of $G$.

  \paragraph{Corrolaries}
  \begin{itemize}[leftmargin=0.45in]
    \item[\theorem{11.2}] The order of every element of $G$ is a divisor of the
      order of $G$.

    \item[\theorem{11.3}] If $G$ has prime order, then $G$ is cyclic.

    \item[\theorem{11.4}] If $x$ is an element of $G$ then $x^{|G|}=e$.

    \item[\theorem{11.5}] \textbf{Euler's Theorem.} If the highest common
      factor of $x$ and $n$ is $1$, then $x^{\phi{(n)}}$ is congruent to $1$
      modulo $n$.

    \item[\theorem{11.6}] \textbf{Fermat's Little Theorem.} If $p$ is prime and
      if $x$ is not a multiple of $p$, then $x^{p-1}$ is congruent to $1$
      modulo $p$.
  \end{itemize}

  \section{Partitions}
  Let $X$ be a set and let $\mathscr{R}$ be a subset of the cartesian product
  $X\times X$. Given two points $x$ and $y$ of $X$, we say that $x$ is
  \textbf{\textit{related}} to $y$ if the ordered pair $(x,y)$ happen to lie in
  $\mathscr{R}$. If (a) each $x\in X$ is related to itself, (b) if $x$ is related
  to $y$, then $y$ is related to $x$, for any two points $x,y \in X$, (c) if
  $x$ is related to $y$ and if $y$ is related to $z$, then $x$ is related to
  $z$ for any three points $x,y,z \in X$ --- then we call $\mathscr{R}$ and
  \textbf{\textit{equivalence relation}} on $X$. For each $x\in X$ the
  collection of all points which are related to it is written $\mathscr{R}(x)$
  and called the \textbf{\textit{equivalence class}} of $x$.

  \paragraph{Theorems}
  \begin{itemize}[leftmargin=0.45in]
    \item[\theorem{12.1}] $\mathscr{R}(x) = \mathscr{R}(y)$ whenever $(x,y) \in
      \mathscr{R}$.
  \end{itemize}

\end{document}
